---
title: Exploring weather related health and economic consequences in the US  
author: "Coursera Student"
date: "September 18, 2015"
output: html_document
---

### Synopsis

This report analysies the health and economic impact of weather related events in the US. This analysis is based on a database of weather events maintained by National Weather Service. This report uses data starting from 1995, although the complete database tracks events as far back as 1950. To limit the scope of the analysis, a reasonable attempt is made to map the 900+ weather events in the database to a more managable 48 events National defined in Weather Service Storm Data Documentation.

Summarise the analysis and conclusion here. Blah Blah Blah Blah Blah Blah Blah Blah Blah Blah Blah Blah Blah Blah BlahBlah Blah Blah Blah BlahBlah Blah Blah Blah Blah Blah Blah Blah Blah Blah Blah Blah Blah Blah Blah Blah Blah Blah Blah BlahBlah Blah Blah Blah BlahBlah Blah Blah Blah Blah

Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

Across the United States, which types of events have the greatest economic consequences?

### Data Processing
There should be a section titled Data Processing which describes (in words and code) how the data were loaded into R and processed for analysis. In particular, your analysis must start from the raw CSV file containing the data. You cannot do any preprocessing outside the document. If preprocessing is time-consuming you may consider using the cache = TRUE option for certain code chunks.
#
Prepare R Environment - load required pacakges and document the session information for reference
#
```{r}
req.pkg <- c("ggplot2","dplyr","lubridate","tidyr","dygraphs")
pkgs.not.installed <- req.pkg[!sapply(req.pkg, function(p) require(p, character.only=T))]
if (length(pkgs.not.installed)>0) install.packages(pkgs.not.installed, dependencies=TRUE)
# load required packages
library(ggplot2)
library(dplyr)
library(tidyr)
library(lubridate)
library(dygraphs)
```
#
####Read the storm database CSV into R data frame
#
```{r}
FSData <- read.csv("./data/repdata-data-StormData.csv")

```

Use the National Weather Service Storm Data Documentation as reference to 'official' set of 48 weather events that will be used for this analysis. As best as possible, map all weather events in the raw dataset this set of 48 weather events.Refer to page 6 Table 1. Storm Data Event Table in [link](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) 


```{r}
# Map to official weather events
EVTYPE <- c("Astronomical Low Tide", "Avalanche", "Blizzard", "Coastal Flood", "Cold/Wind Chill", "Debris Flow", "Dense Fog", "Dense Smoke", "Drought", "Dust Devil", "Dust Storm", "Excessive Heat", "Extreme Cold/Wind Chill", "Flash Flood", "Flood", "Freezing Fog", "Frost/Freeze", "Funnel Cloud", "Hail", "Heat", "Heavy Rain", "Heavy Snow", "High Surf", "High Wind", "Hurricane/Typhoon", "Ice Storm", "Lakeshore Flood", "Lake-Effect Snow", "Lightning", "Marine Hail", "Marine High Wind", "Marine Strong Wind", "Marine Thunderstorm Wind", "Rip Current", "Seiche", "Sleet", "Storm Tide", "Strong Wind", "Thunderstorm Wind", "Tornado", "Tropical Depression", "Tropical Storm", "Tsunami", "Volcanic Ash", "Waterspout", "Wildfire", "Winter Storm", "Winter Weather")
EVTYPE <- as.data.frame(EVTYPE)

EVTYPE$EVTYPE <- tolower(EVTYPE$EVTYPE) # convert to lower case
EVTYPE$EVTYPE101605 <- EVTYPE$EVTYPE 
```

This report uses most recent data from 1995 onwards. It is the majority of the data collected by the National Weather Service Storm Data and is most representative and relevant to planning by government officials responsible for allocating resources to mitigate severe weather events.
```{r}
FSData <- FSData %>% select (BGN_DATE,EVTYPE,FATALITIES,INJURIES,PROPDMG,CROPDMG)
FSData$BGN_DATE <- as.Date(FSData$BGN_DATE,format = "%m/%d/%Y")
FSData$year <- year(FSData$BGN_DATE)
FSData$decade <- paste(substr(FSData$year,1,3),"0's",sep="")
FSData$year <- as.factor(FSData$year)
p <- qplot(decade, data=FSData, geom="histogram",binwidth=20)
p

```

####Health Impact - prepare data set, assign 'official' event type

```{r}
# select only records with FATALITIES,INJURIES
FSDataHealth <-  FSData %>% 
                    select (BGN_DATE,EVTYPE,FATALITIES,INJURIES,year,decade) %>%
                         filter (BGN_DATE >= "1995-01-01", (FATALITIES + INJURIES) > 0 )
nrow(FSDataHealth)
FSDataHealth$EVTYPE <- tolower(FSDataHealth$EVTYPE)

#p <- qplot(year, data=FSDataHealth, geom="histogram",binwidth=20)
#p
```

####Health Impact - Assign official event types 

```{r}
# create raw event list
rawEvents <- unique(FSDataHealth$EVTYPE)
rawEvents <- as.data.frame(rawEvents)
rawEvents$EVTYPE <- tolower(rawEvents$rawEvents) # convert to lower case
rawEvents <- left_join(rawEvents, EVTYPE, by = "EVTYPE")
# sort by event type
rawEvents <- arrange(rawEvents, EVTYPE)
# change to character so DPLYR left_join can function
FSDataHealth$EVTYPE <- as.character(FSDataHealth$EVTYPE)
# sort by event type
FSDataHealth <- arrange(FSDataHealth, EVTYPE)
FSDataHealth <- left_join(FSDataHealth, rawEvents, by = "EVTYPE")
# change back to factor
FSDataHealth$EVTYPE <- as.factor(FSDataHealth$EVTYPE)
FSDataHealth$EVTYPE101605 <- as.factor(FSDataHealth$EVTYPE101605)
#p <- qplot(is.na(EVTYPE101605), data=FSDataHealth, geom="histogram",binwidth=20)
#p

```

####Health Impact - Analyse and report results 

```{r}
FSDataHealthFinal <-  FSDataHealth %>% 
    select (year,EVTYPE101605,FATALITIES,INJURIES)  %>%
        group_by(year,EVTYPE101605) %>%
            filter(!is.na(EVTYPE101605)) %>%
                mutate(impact= FATALITIES+INJURIES)

FSDataHealthFinal$FATALITIES <- NULL
FSDataHealthFinal$INJURIES   <- NULL
## FSDataHealthFinal$year   <- NULL

FSDataHealthFinal <- FSDataHealthFinal %>%
                        select (year,EVTYPE101605,impact) %>%
                            group_by (year,EVTYPE101605) %>%
                              summarise(impact=sum(impact))

FSDataHealthFinal <- FSDataHealthFinal %>%
                        select (year,EVTYPE101605,impact) %>%
                            group_by (EVTYPE101605) %>%
                              summarise(healthimpact=sum(impact))

FSDataHealthFinal <- FSDataHealthFinal %>%
                            arrange(desc(healthimpact))

FSDataHealthFinal <- head(FSDataHealthFinal,10) # select top 10

g <- ggplot(FSDataHealthFinal, aes(EVTYPE101605, healthimpact)) 
g + geom_bar(stat = "identity")
```


####Economic Impact - prepare data set, assign official event type

```{r}
# select only records with PROPDMG,CROPDMG
FSDataEconomic <- FSData %>% select (BGN_DATE,EVTYPE,PROPDMG,CROPDMG)
FSDataEconomic$BGN_DATE <- as.Date(FSDataEconomic$BGN_DATE,format = "%m/%d/%Y")
FSDataEconomic$year <- year(FSDataEconomic$BGN_DATE)
FSDataEconomic$decade <- paste(substr(FSDataEconomic$year,1,3),"0's",sep="")
FSDataEconomic$year <- as.factor(FSDataEconomic$year)
p <- qplot(decade, data=FSDataEconomic, geom="histogram",binwidth=20)
p
FSDataEconomic <-  FSDataEconomic %>% 
                            select (BGN_DATE,EVTYPE,PROPDMG ,CROPDMG,year,decade) %>%
                         filter (BGN_DATE >= "1995-01-01", (PROPDMG + CROPDMG) > 0 )
FSDataEconomic$EVTYPE <- tolower(FSDataEconomic$EVTYPE)
#p <- qplot(year, data=FSDataEconomic, geom="histogram",binwidth=20)
#p

```

####Economic Impact - Assign official event types 

```{r}

# create raw event list
rawEvents <- unique(FSDataEconomic$EVTYPE)
rawEvents <- as.data.frame(rawEvents)
rawEvents$EVTYPE <- tolower(rawEvents$rawEvents) # convert to lower case
rawEvents <- left_join(rawEvents, EVTYPE, by = "EVTYPE")
# sort by event type
rawEvents <- arrange(rawEvents, EVTYPE)
# change to character so DPLYR left_join can function
FSDataEconomic$EVTYPE <- as.character(FSDataEconomic$EVTYPE)
# sort by event type
FSDataEconomic<- arrange(FSDataEconomic, EVTYPE)
FSDataEconomic <- left_join(FSDataEconomic, rawEvents, by = "EVTYPE")
# change back to factor
FSDataEconomic$EVTYPE <- as.factor(FSDataEconomic$EVTYPE)
FSDataEconomic$EVTYPE101605 <- as.factor(FSDataEconomic$EVTYPE101605)
#p <- qplot(is.na(EVTYPE101605), data=FSDataEconomic, geom="histogram",binwidth=20)
# p
```

####Economic Impact - Analyse and report results

```{r}
FSDataEconomicFinal <-  FSDataEconomic %>% 
    select (year,EVTYPE101605,PROPDMG,CROPDMG)  %>%
        group_by(year,EVTYPE101605) %>%
            filter(!is.na(EVTYPE101605)) %>%
                mutate(impact= PROPDMG+CROPDMG)

FSDataEconomicFinal$PROPDMG <- NULL
FSDataEconomicFinal$CROPDMG   <- NULL
## FSDataEconomicFinal$year   <- NULL

FSDataEconomicFinal <- FSDataEconomicFinal %>%
                        select (year,EVTYPE101605,impact) %>%
                            group_by (year,EVTYPE101605) %>%
                              summarise(impact=sum(impact))

FSDataEconomicFinal <- FSDataEconomicFinal %>%
                        select (year,EVTYPE101605,impact) %>%
                            group_by (EVTYPE101605) %>%
                              summarise(econimpact=sum(impact))

FSDataEconomicFinal <- FSDataEconomicFinal %>%
                            arrange(desc(econimpact))

FSDataEconomicFinal <- head(FSDataEconomicFinal,10) # select top 10

g <- ggplot(FSDataEconomicFinal, aes(EVTYPE101605, econimpact)) 
g + geom_bar(stat = "identity")
```

### Results

There should be a section titled Results in which your results are presented.You may have other sections in your analysis, but Data Processing and Results are required. The analysis document must have at least one figure containing a plot. Your analyis must have no more than three figures. Figures may have multiple plots in them (i.e. panel plots), but there cannot be more than three figures total.You must show all your code for the work in your analysis document. This may make the documenta bit verbose, but that is okay. In general, you should ensure that echo = TRUE for every code chunk (this is the default setting in knitr).
